Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-05 paper/main.tex
Updates

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations}
\todo[inline]{Improve this section once all the notations are in place}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $, conditional
expectations as $ \mathbb{E}[X | \bm{Z}] $.

\subsection{Measures of Conditional Association}
Our proposed method for iterative DAG construction relies on a measure of
conditional association also known as partial association between variables. We
are interested in measuring the association between variables $ X $ and $ Y $
when conditioned on a set of variables $ \bm{Z} $ which is potentially an empty
set, $ \bm{Z} = \emptyset $. Many special case measures of conditional
association have been used in literature depending on the type of $ X $, $ Y $,
and $ \bm{Z} $. Typically, these are the effect size measures of the CI test $
X \ci Y \rvert \bm{Z} $. In this section, we give an overview of some of the
commonly used measures for different data types.

@@

where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively. When $ \bm{Z} \neq \emptyset $, partial correlation coefficient
can be used. One of the ways to estimate it is to use two linear regression
models:

\begin{equation}
	\begin{split}
		E_X &: X \sim \bm{Z} \\
		E_Y &: Y \sim \bm{Z} \\
		\rho_{X, Y; \bm{Z}} &= \frac{cov((X - E_X), (Y - E_Y))}{\sigma_{(X - E_X)} \sigma_{(Y-E_Y)}}
	\end{split}
\end{equation}

$ E_X: x \sim \bm{z} $, and $ E_Y: y \sim \bm{z} $. Then we take the compute
the Pearson's correlation coefficient between the residuals of these two
regressions: $ R_X = x - E_X(\bm{z}) $, $ R_Y = y - E_Y(\bm{z}) $.

@@

In the case of discrete variables, Cramer\'s V has been typically used as a
measure of association. When there are no conditional variables, $ \bm{Z} =
\emptyset $, Cramer\'s V can be computed directly from the contingency table.

However, when conditional variables are present, we can iterate over all
possible combinations of the conditional variables, compute the Cramer\'s V and
then lastly combine them together.

\paragraph{$ X $ is ordinal and $ Y $ is continuous or ordinal}
Polyserial Correlation (when one is oridinal and the other is continuous) and
polychoric correlation (when both are ordinal) can be use. Here the assumption
is made that $ X $ is a thresholded version of an underlying continuous normal
variable. The method then tries to learn the thresholds for this latent
variable while trying to maximize the correlation between $ X_{lat} $ and $ Y
$. This gives us values that are Pearson's correlation coefficient.

@@

\begin{enumerate}
	\item \textbf{Bounded: } $ \rho $ is bounded between $ 0 $ and $ 1 $,
		with $ 0 $ representing no association while $ 1 $ represents
		perfect linear association. This helps in interpretation and
		comparing values across multiple partial associations.
	\item \textbf{Independent of Sample Size: } The measure remains consistent
		regardless of the sample size ensuring reliability across 
		studies with different sample sizes.
	\item \textbf{Invariant to the Dummy Encoding Scheme: } Since canonical
		correlations search for linear combinations over the columns,
		the measure is invariant to the specific dummy encoding
		scheme used for categorical variables.
	\item \textbf{Equivalent to Partial Correlations for Continuous variables}
		Canonical correlations is equivalent to Pearson's correlation coefficient
		when both $ X $ and $ Y $ are continuous. Hence, $ \rho $ generalizes
		partial correlations to mixed data types.
	\item \textbf{Equivalent to polychoric and polyserial correlation for ordinal and continuous variables: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cramer\'s V: } 
\end{enumerate}

>>>>

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations}
\todo[inline]{Improve this section once all the notations are in place}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association.

\subsection{Measures of Conditional Association}
Our proposed method for iterative DAG construction relies on a measure of
conditional association, also commonly known as partial association, between
variables. Specifically, we are interested in quantifying the association
between variables $ X $ and $ Y $ when conditioned on a set of variables $
\bm{Z} $, which may be empty ($ \bm{Z} = \emptyset $). Various measures of
conditional association have been used in the literature, depending on the type
of $ X $, $ Y $, and $ \bm{Z} $. These measures are often based on the effect
size of the CI test $ X \ci Y \rvert \bm{Z} $. In this section, we provide an
overview of some of the commonly used measures for different data types.

@@

where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively. When $ \bm{Z} \neq \emptyset $, partial correlation coefficient
can be used. One of the ways to estimate it is to use two regression
models $ E_X: X \sim \bm{Z} $, $ E_Y: Y \sim \bm{Z} $, and compute the Pearson's 
correlation coefficient between the residuals of these regression models:

\begin{equation}
	\begin{split}
		R_X &= X - E_X(\bm{Z})  \;\;\;\;\;\;\;\;  R_Y = Y - E_Y(\bm{Z}) \\
		\rho_{X, Y; \bm{Z}} &= \frac{\mathrm{cov}(R_X, R_Y)}{\sigma_{R_X} \sigma_{R_Y}}
	\end{split}
\end{equation}

@@

In the case of all discrete variables, Cram\'er's V can used as a measure of
association. When $ \bm{Z} = \emptyset $, Cram\'er's V can be computed from the 
chi-squared test statistic as:

\begin{equation}
	\mathbb{V_{X, Y}} = \sqrt{\frac{\chi^2 / n}{\mathrm{min}(k-1, r-1)}}
\end{equation}

where $ \chi^2 $ is the chi-squared test statistic for $ X $ and $ Y $, $ n $ 
is the sample size and $ k $ and $ r $ are the number of categories of $ X $ and 
$ Y $.

However, when $ \bm{Z} \neq \emptyset $, we first split the dataset by unique
values of $ \bm{Z} $, $ D = \{D_{\bm{Z} = \bm{Z_1}}, D_{\bm{Z} = \bm{Z_2}},
\cdots, \} $  and then combine them as:

\begin{equation}
	\mathbb{V_{X, Y; Z}} = \sum_{D_{\bm{Z} = \bm{Z_i}}} \mathbb{V}_{X_{D_i}, Y_{D_i}}
\end{equation}

\paragraph{$ X $ is ordinal and $ Y $ is continuous or ordinal}
When one of the variables is ordinal and the other is continuous polyserial
correlation can be used. When both the variables are ordinal polychoric
correlation can be used. Both polychoric and polyserial correlation make the
assumption that the ordinal variable has an underlying normal distributed
continuous variable, and the ordinal variable has been generated by
thresholding this latent variable. The method then tries to maximize the
likelihood of data given polychoric or polyserial correlation coefficient. This
method gives us values that are Pearson's correlation coefficient.

@@

\begin{enumerate}
	\item \textbf{Bounded: } $ \rho $ is bounded between $ 0 $ and $ 1 $,
		with $ 0 $ representing no association while $ 1 $ represents
		perfect linear association. This helps in interpretation and
		comparing values across multiple partial associations.
	\item \textbf{Independent of Sample Size: } The measure remains consistent
		regardless of the sample size ensuring reliability across 
		studies with different sample sizes.
	\item \textbf{Invariant to the Dummy Encoding Scheme: } Since canonical
		correlations search for linear combinations over the columns,
		the measure is invariant to the specific dummy encoding
		scheme used for categorical variables.
	\item \textbf{Equivalent to Partial Correlations for Continuous variables}
		Canonical correlations is equivalent to Pearson's correlation coefficient
		when both $ X $ and $ Y $ are continuous. Hence, $ \rho $ generalizes
		partial correlations to mixed data types.
	\item \textbf{Equivalent to polychoric and polyserial correlation for ordinal and continuous variables: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cram\'ers V: } 
\end{enumerate}
